Designing Types for R, Empirically

### 2020-05-16 paper/main.tex

\noindent
Assuming that type sequences can be reordered freely, we rewrite types to
minimize their size by removing redundant types, types that are subsumed by
subtyping, and \null types. For performance reasons, we bound lists to no
more than five types in their union, above that they become \LIST{any}.
Functions are alos limited to twenty arguments.

@@

In terms of usability, our checker is enabled automatically whenever a
package is loaded in R. So, a simple invocation of \code{library(contractr)}
causes contracts to be injected in all packages.  On loading, \contractr
scans all packages in the user's workspace and inserts contracts in
functions for which type signatures are available. Then \contractr sets up
package load hooks which are executed when new packages are loaded.
Furthermore, \contractr automatically removes contracts from all functions
and restores them to their original state when it is unloaded.

The type signatures can be provided externally, in a file, thus avoiding the
need to change the source code of checked packages.  Type declarations can
also be written alongside a \code{@type} section inside \roxygen function
comment blocks. \roxygen is an R package that enables authors to add
documentation. The documentation sections are processed by \roxygen before
package build and a registered method for each custom tag section is
invoked with the section data and documented object. \contractr uses this
mechanism to register a hook for a custom \code{@type} tag.

The \contractr API allows users to explicitly insert contracts during
interactive development by supplying the type signature as a string.
Contracts can be removed by calling \code{remove\_contract}.  Contracts can
be selectively enabled or disabled for code blocks by wrapping them in calls
to \code{ignore\_contracts} and \code{capture\_contracts}.  These functions
return a data frame that contains all the information about failed and
successful contract assertions in the wrapped code blocks.  Multiple
failures can introduce noise; this is avoided by setting
\code{contractr.severity} to \code{'silence'}.  \contractr still performs
type-checking so that results can be obtained by calling
\code{get\_contracts}. Setting \code{contractr.severity} flag to
\code{'error'} turns the type-checking warnings to errors and
halt the program.

\paragraph{Checking the values} A number of properties can be checked
by a simple tag check, namely whether a value is a: null, environment, vector
or scalar. Other properties require an inspection of the contents of a
value: the absence of \NA or the type of the elements of list. Union types
require checking for each of the alternatives. We do not check higher-order
functions in this version of the tool.

When argument values are wrapped in promises (this is not always the case
due to compiler optimizations), in order to retain the non-strict semantics
of R, the expression held in the promise is wrapped in a call to the type
checker, and type checking is delayed until the promise is forced. This
leads to corner cases such that the type checking of a function may happen
after that function has returned.

Return values require care as well.  Functions return the last expression
they evaluate, thus to avoid having to analyze the code of the called
function the checker will register a callback on the exit hook. The hook is
executed in the function call's environment. Another wrinkle is due to
longjumps which causes active function calls on the stack to be
discarded. When they are discarded, their exit hooks are called.  But, these
function do not have a return value to type-check. \contractr deals with
this problem by allocating a unique sentinel object which serves as the
return value for calls that are discarded. The exit hook does not call the
type-checker if it see the sentinel.

@@

@@

>>>>

\noindent
Assuming that type sequences can be reordered freely, we rewrite types to
minimize their size by removing redundant types, types that are subsumed by
subtyping, and \null types. For performance reasons, we bound lists to no
more than five types in their union, above that they become \LIST{any}.
Functions are alos limited to twenty arguments. Higher-order functions are
approximated by \any to \any.

@@

In terms of usability, our checker is enabled automatically whenever a
package is loaded in R. So, a simple invocation of \code{library(contractr)}
causes contracts to be injected in all packages.  On loading, \contractr
scans all packages in the user's workspace and inserts contracts in
functions for which type signatures are available. Then \contractr sets up
package load hooks which are executed when new packages are loaded.
Furthermore, \contractr automatically removes contracts from all functions
and restores them to their original state when it is unloaded.
The type signatures can be provided externally, in a file, thus avoiding the
need to change the source code of checked packages.  Type declarations can
also be written alongside a \code{@type} section inside \roxygen function
comment blocks. \roxygen is an R package that enables authors to add
documentation. The documentation sections are processed by \roxygen before
package build and a registered method for each custom tag section is
invoked with the section data and documented object. \contractr uses this
mechanism to register a hook for a custom \code{@type} tag. 
The \contractr API allows users to explicitly insert contracts during
interactive development by supplying the type signature as a string.
Contracts can be removed by calling \code{remove\_contract}.  Contracts can
be selectively enabled or disabled for code blocks by wrapping them in calls
to \code{ignore\_contracts} and \code{capture\_contracts}.  These functions
return a data frame that contains all the information about failed and
successful contract assertions in the wrapped code blocks.  Multiple
failures can introduce noise; this is avoided by setting
\code{contractr.severity} to \code{'silence'}.  \contractr still performs
type-checking so that results can be obtained by calling
\code{get\_contracts}. Setting \code{contractr.severity} flag to
\code{'error'} turns the type-checking warnings to errors and
halt the program.

\paragraph{Checking the values} A number of properties can be checked
by a simple tag check, namely whether a value is a: null, environment, vector
or scalar. Other properties require an inspection of the contents of a
value: the absence of \NA or the type of the elements of list. Union types
require checking for each of the alternatives. We do not check higher-order
functions in this version of the tool.
When argument values are wrapped in promises (this is not always the case
due to compiler optimizations), in order to retain the non-strict semantics
of R, the expression held in the promise is wrapped in a call to the type
checker, and type checking is delayed until the promise is forced. This
leads to corner cases such that the type checking of a function may happen
after that function has returned.
Return values require care as well.  Functions return the last expression
they evaluate, thus to avoid having to analyze the code of the called
function the checker will register a callback on the exit hook. The hook is
executed in the function call's environment. Another wrinkle is due to
longjumps which causes active function calls on the stack to be
discarded. When they are discarded, their exit hooks are called.  But, these
function do not have a return value to type-check. \contractr deals with
this problem by allocating a unique sentinel object which serves as the
return value for calls that are discarded. The exit hook does not call the
type-checker if it see the sentinel.

@@

We ran \typetracer on the test, example, and vignette code of the aforementioned corpus of \PACKAGES packages and inferred types for \NUMFUNCTIONS functions.
We have selected and will discuss a few interesting signatures.

@@

Many of the features of our type language are represented here.
Some of these signatures are telling of the function's behaviour.
For example, consider the signature for \code{openssl::decrypt_envelope}: the firs three parameters of the function are byte arrays, and the fourth argument is an RSA key, likely one used to decrypt some of the inputs, and the output of the function is another byte array.
As another example, consider \code{data.tree::Traverse}: according to the function documentation, it takes the root of a tree and traverses it in an order specified by the second argument.
